Implementing Distributed Systems Directly with I/O Automata in Ioa++

### 2011-09-13 papers/paper1/component_model.tex
Changed tcp_lcr to use asynch_lcr_automaton.

In this section we describe how I/O automata offer a natural realization of components, interfaces, and provide natural support for concurrency and composition.
The Input/Output (I/O) automata model was developed by Lynch~\cite{lynch1996distributed} to model reactive concurrent and asynchronous systems.
An I/O automaton consists of state variables and a set of atomic actions that manipulate the state variables.
An action is only permitted to manipulate the state variables of the automaton to which it belongs.
An action is either an internal action, an output action, or an input action.
An internal action changes the state of the automaton.
An output action changes the state of the automaton and produces a signal or value that is consumed by zero or more input actions.
An input action changes the state of the automaton when it receives a signal or value produced by an output action.
\emph{Unvalued} output and input actions produce and consume signals respectively.
\emph{Valued} output and input actions produce and consume values respectively.
The set of actions associated with an automaton is known as the automaton's \emph{signature}~\cite{lynch1996distributed}.

An automaton's output and internal actions constitute its \emph{local signature}~\cite{lynch1996distributed}.
Often, local actions consists of a precondition and an effect.
The precondition is a predicate over the state variables that enables the effect when selected.
The effect is a function that computes the next state of the automaton from the current state.
Input actions presented in this style consist solely of an effect.

@@

Execution in the I/O automata model consists of non-deterministically and repeatedly selecting a local action and then applying the effect if the precondition is true.
To be general, the scheduler is assumed to be fair meaning that a local action is guaranteed to be selected (but not necessarily executed) infinitely often.
Executing internal actions and output actions that are not composed with any input actions is straightforward: in one atomic step, the precondition is evaluated and the effect is applied if the precondition is true.
Executing a composed output action consists of atomically evaluating the precondition and applying the effect of the output action and all composed input actions using the value produced by the output action subject to the precondition.
The atomic relationship between composed output actions and input actions means that multiple automata may change state atomically when executing an output action.
The guarantee of atomicity allows a developer to relate the state of one automaton to another, resulting in well-defined interactions that facilitate composition.

@@

As a mathematical abstraction the I/O automata model is appropriate for modeling reactive asynchronous and concurrent systems.
However, it lacks certain features needed to develop real components.
This subsection introduces these features and lays the foundation for the dynamics discussed in section~\ref{dynamics}.

\paragraph{Recursive Encapsulation.}
Component frameworks that support recursive encapsulation allow components to contain other components.
Recursive encapsulation encodes the idea of ``divide-and-conquer'' and is at the heart of both bottom-up and top-down design strategies.
The contained component is the \emph{child} while the containing component is the \emph{parent}.
The resulting hierarchy helps developers organize programs by separating concerns and facilitates the reuse of existing components.
A distinction is made between the \emph{name} given to an instance and the \emph{type} of a child component to allow a parent component to contain multiple children of the same type.

Recursive encapsulation means that an I/O automaton may contain and use other I/O automata.
Recursive encapsulation implies that a parent automaton is the result of composing its children and an anonymous \emph{kernel automaton}.
The kernel automaton allows a parent automaton to have its own state variables and actions.
The \emph{root automaton} is the automaton at the top of the hierarchy and represents the entire system as composition is applied recursively up the hierarchy.

Figure\ref{sys_model} shows a sample event detector system where automata are depicted as solid rectangles, e.g., Processor.
Named automata are labeled by a box in their upper left corner, e.g., C.
The sample event detector system illustrated in Figure~\ref{sys_model} contains three automata (components):  an anonymous Processor, a Converter named C, and a Detector named D.
The Processor automaton is the root and C and D are its children.

@@

\paragraph{Binding rules.}
To be compliant with name-based composition under the I/O automata model, the set of bindings must adhere to certain rules.
First, the output action and input action of a binding must agree on the type being produced and consumed.
For example, an input action consuming an integer value can only be bound to an output action that produces an integer value.
The system shown in figure~\ref{sys_model} follows this rule since $(out!(s), C.in?(s))$ agree on type $s$, $(C.out!(t), D.in?(t))$ agree on type $t$, and $(D.detect!, detected?)$ agree on the absence of a value (i.e., a signal).
Second, an input action can be bound to at most one output action.
Composing an input action with multiple output actions would introduce ambiguity into the relationship between the states of the automata and is therefore prohibited by the model.
A binding map where all bindings that mention the same input action are equivalent satisfies this rule.
For dynamics, which are discussed in section~\ref{dynamics}, we must strengthen this rule to say that an input action can appear at most once in the global binding map.
Consequently, every binding has exactly one owner.
For a graphical representation such as Figure~\ref{sys_model}, this rule means that an input is the sink of at most one arrow (binding).
Third, an output action cannot be bound to more than one input action in the same automaton:  the state of the automaton containing the input actions would not be well-defined after the output is executed since the input action effects would be applied in an undefined order.
Graphically, arrows originating at the same output action terminate in different components.
%% Let $\pi$ be a prefix operator that strips the action name (leaving the path and hence the automaton) of a fully-qualified action name.
%% The following must be true for the global binding map $B$:  $\langle \forall o : (o, z) \in B :: \langle \forall i,j : (o, i) \in B \land (o, j) \in B \land i \neq j:: \pi (i) \neq \pi (j) \rangle \rangle$.
Fourth, an output action cannot be bound to an input action in the same automaton.
Allowing such bindings would result in undefined behavior since the order of the output and input effects is undefined.
The I/O automata model prevents this by requiring that all actions belonging to the same automaton have a unique name.
To elaborate, we usually think about the output action occurring before the input action because a value must be generated.
However, the I/O automata model admits an interpretation where the value is generated first and the effects of the output action and input action are applied non-deterministically.
Graphically, an arrow originating in one component must terminate in another component.
%% To adhere to this rule, the following must be true for the global binding map $B$:  $\langle \forall o, i : (o, i) \in B :: \pi (o) \neq \pi (i) \rangle$.

>>>>

In this section we describe how I/O automata offer a natural realization of components and interfaces, and provide natural support for concurrency and composition.
The Input/Output (I/O) automata model was developed by Lynch~\cite{lynch1996distributed} to model reactive concurrent and asynchronous systems.
An I/O automaton consists of state variables and a set of atomic actions that manipulate the state variables.
An action is only permitted to manipulate the state variables of the automaton to which it belongs.
An action is either an internal action, an output action, or an input action.
An internal action changes the state of the automaton.
An output action changes the state of the automaton and produces a signal or value that is consumed by zero or more input actions.
An input action changes the state of the automaton when it receives a signal or value produced by an output action.
\emph{Unvalued} output and input actions produce and consume signals respectively.
\emph{Valued} output and input actions produce and consume values respectively.
The set of actions associated with an automaton is known as the automaton's \emph{signature}~\cite{lynch1996distributed}.

An automaton's output and internal actions constitute its \emph{local signature}~\cite{lynch1996distributed}.
Often, a local action consists of a precondition and an effect.
The precondition is a predicate over the state variables that enables the effect when selected.
The effect is a function that computes the next state of the automaton from the current state.
An input action presented in this style consists solely of an effect.

@@

Execution in the I/O automata model consists of non-deterministically and repeatedly selecting a local action and then applying the effect if the precondition is true.
In general the scheduler is assumed to be fair, meaning that a local action is guaranteed to be selected (but not necessarily executed) infinitely often.
Executing internal actions and output actions that are not composed with any input actions is straightforward: in one atomic step, the precondition is evaluated and the effect is applied if the precondition is true.
Executing a composed output action consists of atomically evaluating the precondition and applying the effect of the output action and all composed input actions, using the value produced by the output action subject to the precondition.
The atomic relationship between composed output actions and input actions means that multiple automata may change state atomically when executing an output action.
The guarantee of atomicity allows a developer to relate the state of one automaton to another, resulting in well-defined interactions that facilitate composition.

@@

As a mathematical abstraction the I/O automata model is appropriate for modeling reactive asynchronous and concurrent systems.
However, it lacks certain features needed to develop real components.
This subsection introduces these features and lays the foundation for adding the dynamics discussed in section~\ref{dynamics}.

\paragraph{Recursive Encapsulation.}
Component frameworks that support recursive encapsulation allow components to contain other components.
Recursive encapsulation is at the heart of both bottom-up and top-down design strategies.
The contained component is the \emph{child} while the containing component is the \emph{parent}.
The resulting hierarchy helps developers organize programs by separating concerns and facilitates the reuse of existing components.
A distinction is made between the \emph{name} given to an instance and the \emph{type} of a child component to allow a parent component to contain multiple children of the same type.

In ioa++, recursive encapsulation means that an I/O automaton may contain and use other I/O automata.
Recursive encapsulation implies that a parent automaton is the result of composing its children and an anonymous \emph{kernel automaton}.
The kernel automaton allows a parent automaton to have its own state variables and actions.
The \emph{root automaton} is the automaton at the top of the hierarchy and represents the entire system as composition is applied recursively up the hierarchy.

Figure~\ref{sys_model} shows a sample event detector system where automata are depicted as solid rectangles, e.g., Processor.
Named automata are labeled by a box in their upper left corner, e.g., C.
The sample event detector system illustrated in Figure~\ref{sys_model} contains three automata (components):  an anonymous Processor, a Converter named C, and a Detector named D.
The Processor automaton is the root and C and D are its children.

@@

\paragraph{Binding rules.}
To be compliant with name-based composition under the I/O automata model, the set of bindings must adhere to certain rules.
First, the output action and input action of a binding must agree on the type being produced and consumed.
For example, an input action consuming an integer value can only be bound to an output action that produces an integer value.
The system shown in Figure~\ref{sys_model} follows this rule since $(out!(s), C.in?(s))$ agree on type $s$, $(C.out!(t), D.in?(t))$ agree on type $t$, and $(D.detect!, detected?)$ agree on the absence of a value (i.e., a signal).
Second, an input action can be bound to at most one output action.
Composing an input action with multiple output actions would introduce ambiguity into the relationship between the states of the automata and is therefore prohibited by the model.
A binding map where all bindings that mention the same input action are equivalent satisfies this rule.
For dynamics, which are discussed in section~\ref{dynamics}, we must strengthen this rule to say that an input action can appear at most once in the global binding map.
Consequently, every binding has exactly one owner.
For a graphical representation such as Figure~\ref{sys_model}, this rule means that an input is the sink of at most one arrow (binding).
Third, an output action cannot be bound to more than one input action in the same automaton:  the state of the automaton containing the input actions would not be well-defined after the output is executed since the input action effects would be applied in an undefined order.
Graphically, arrows originating at the same output action terminate in different components.
%% Let $\pi$ be a prefix operator that strips the action name (leaving the path and hence the automaton) of a fully-qualified action name.
%% The following must be true for the global binding map $B$:  $\langle \forall o : (o, z) \in B :: \langle \forall i,j : (o, i) \in B \land (o, j) \in B \land i \neq j:: \pi (i) \neq \pi (j) \rangle \rangle$.
Fourth, an output action cannot be bound to an input action in the same automaton.
Allowing such bindings would result in undefined behavior since the order of the output and input effects is undefined.
The I/O automata model prevents this by requiring that all actions belonging to the same automaton have a unique name.
To elaborate, we usually think about the output action occurring before the input action because a value must be generated.
However, the I/O automata model admits an interpretation where the value is generated first and the effects of the output action and input action are applied non-deterministically.
Graphically, an arrow originating in one component must terminate in another component.
%% To adhere to this rule, the following must be true for the global binding map $B$:  $\langle \forall o, i : (o, i) \in B :: \pi (o) \neq \pi (i) \rangle$.

### 2011-09-13 papers/paper1/design.tex
Changed tcp_lcr to use asynch_lcr_automaton.

An implementation of the model described in section~\ref{component_model} must provide a service for executing local actions and facilities for manipulating the set of automata and bindings.
The architecture of our framework is depicted in Figure~\ref{framework_architecture}.
The Scheduler has the resposibility of selecting and executing local actions subject to the I/O automata model and concurrency constraints of~\ref{component_model}.
All user automata inherit from a common base class that provides facilities for asynchronously creating and destroying child automata and bindings.
The services provided by the base class are indicated by the Automaton layer of Figure~\ref{framework_architecture}.
The Automaton class contains statically composed System Actions that are used to communicate with the System Automaton.
In most situations, the services provided by the Automaton base class are too low-level to be used directly by a User Automaton.
An Automaton Manager presents a high-level interface for asynchronously managing a single child automaton using the services of the Automaton base class.
Similarly, a Binding Manager presents a high-level interface for asynchronously managing a single binding.
We assume a POSIX environment whose native concurrency mechanism is threads.

@@

\paragraph{Actions of non-existent automata.}
Every action in the set of actions considered by the controller was scheduled by its associated automaton.
Since automata can be destroyed, an action is the set of actions owned by the controller might belong to an automaton that no longer exists.
Our solution is to check if the automaton exists before executing an action.
This check is performed by the model and avoids the overhead associated with synchronizing the set of actions with the set of automata every time an automaton is destroyed.
Our solution assumes that each automaton is assigned a unique identifier and that the set of actions turns over comparatively faster than the set of identifiers.

>>>>

An implementation of the component model described in section~\ref{component_model} must provide a service for executing local actions and facilities for manipulating the set of automata and bindings.
The architecture of our framework is depicted in Figure~\ref{framework_architecture}.
The Scheduler has the resposibility of selecting and executing local actions subject to the I/O automata model and concurrency constraints of~\ref{component_model}.
All user automata inherit from a common base class that provides facilities for asynchronously creating and destroying child automata and bindings.
The services provided by the base class are indicated by the Automaton layer of Figure~\ref{framework_architecture}.
The Automaton class contains statically composed System Actions that are used to communicate with the System Automaton.
In most situations, the services provided by the Automaton base class are too low-level to be used directly by a User Automaton.
An Automaton Manager presents a high-level interface for asynchronously managing a single child automaton using the services of the Automaton base class.
Similarly, a Binding Manager presents a high-level interface for asynchronously managing a single binding.
We assume a POSIX environment whose native concurrency mechanism is threads.

@@

\paragraph{Actions of non-existent automata.}
Every action in the set of actions considered by the controller was scheduled by its associated automaton.
Since automata can be destroyed, an action in the set of actions owned by the controller might belong to an automaton that no longer exists.
Our solution is to check if the automaton exists before executing an action.
This check is performed by the model and avoids the overhead associated with synchronizing the set of actions with the set of automata every time an automaton is destroyed.
Our solution assumes that each automaton is assigned a unique identifier and that the set of actions turns over comparatively faster than the set of identifiers.

### 2011-09-13 papers/paper1/programming_model.tex
Finished tcp_lcr.
Changed tcp_lcr to use asynch_lcr_automaton.

\section{I/O Automata Representation and Programming Model\label{representation}}

In this section, we show how I/O automata are represented using our framework and how they can be used to construct programs.
We develop a simulation for a distributed algorithm to show the basic features of the framework and then show a real implementation of the same algorithm.
The problem we consider is that of electing a leader in unidirectional ring.

The algorithm we consider is the asynchronous LCR algorithm presented by Lynch in Chapter 15 of~\cite{lynch1996distributed} which is in turn based on the work of LeLann~\cite{} and Chang and Roberts~\cite{}.
The LCR algorithm assumes that every node has a unique identifier.
Each node begins by assuming that it is the leader and send its own unique id to its neighbor.
If a node receives a unique id from its predecessor that is greater that the id that it believes to be the leader, then it changes its belief about the leader to the new id and sends the new id to its successor.
When a node receives its own id, it elects itself the leader of the ring.

>>>>

\section{I/O Automata Programming Model\label{representation}}

In this section, we show how I/O automata can be used to construct programs by implementing an algorithm for electing a leader in a unidirectional ring.
The algorithm we consider is the asynchronous LCR algorithm presented by Lynch in Chapter 15 of~\cite{lynch1996distributed} which is in turn based on the work of LeLann~\cite{le1977distributed} and Chang and Roberts~\cite{chang1979improved}.
The LCR algorithm assumes that every node has a unique identifier (UID).
When the protocol begins, each node sends its UID to its successor.
When a node receives a UID from its predecessor that is greater than its own UID, it forwards the UID to its successor.
When a node receives its own UID, it elects itself the leader of the ring.

Automata are classes

Actions (2/3 functions + MACRO)

binding predicate (binding count)

dynamics
